Forecasting the United States 2024 Presidential Election

### 2024-11-03 paper/paper.qmd

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

@@

One particular poll within our data sample is one conducted between October 11-14 by Beacon Research and Shaw & Company Research (cite Fox article). Their analysis found 48% of participants favoured Harris while 50% favoured Trump. Targeting the sample frame of registered American voters, this poll collected data on 1,110 participants through either live phonecall interviews or an online survey. The sample was recruited by applying the probability proportionate to size method on the nationwide voter file of registered voters’ phone numbers, meaning that the contacted individuals would be proportionally representative of the number of voters per state (cite Fox article). This ensures that the voices represented by poll results would match the ones involved with the real election as much as possible. 
=======
## Analysis into a Pollster's Methodology

One particular poll within our data sample is one conducted between October 11-14 by Beacon Research and Shaw & Company Research (cite Fox article). Targeting the sample frame of registered American voters, this poll collected data on 1,110 participants’ responses on 51 questions through either live phonecall interviews or an online survey. The sample was recruited by applying the probability proportionate to size method on the nationwide voter file of registered voters’ phone numbers, meaning that the contacted individuals would be proportionally representative of the number of voters per state (cite Fox article). This ensures that the voices represented by poll results would match the ones involved with the real election as much as possible.

The key finding from this poll is that 48% of participants favored Harris while 50% favored Trump, with the lead being consistent in the larger sample of registered voters and smaller subsample of likely voters. Despite this, responses also indicate that Harris is narrowly leading in seven swing states, meaning that Democrats could potentially win by electoral college while losing the popular vote (cite Fox article). This report also notes that current results are Trump’s highest approval ratings since Biden dropped out of the race, while support for Harris is at its lowest.

One strength of this questionnaire is that information regarding the respondent’s confidence level and commitment level was also collected. With questions such as “How often do you make a point to read or listen to the news?” and “Are you certain to support that candidate, or do you think you may change your mind and support someone else?”, this poll is able to gain a more indepth view of how easily these participants may be swayed in the time between the polling and the real election (cite Fox article). As well, probabilistic statistical models based on past voting history, interest in current election, age, education, race, ethnicity, church attendance, and marital status were used to predict the respondents’ likeliness to vote; interestingly, results found that the results for the full sample vs subsample of 870 likely voters varied by a ±3%, which can be quite significant given how tight the race is currently (cite Fox article).

On the other hand, one potential weakness of this poll is that it was sponsored by Fox News, a news site that has historically been Republican-leaning. Though from the reported methodology alone, no evident biasing of the pollster can be observed, this potential source of biased reporting must be noted as it can affect the type of analyses that occurred and which key findings are the focus of news reports. As well, another potential issue is the lack of indication regarding how non-responses of “Don’t know” were handled during analysis (cite Fox article). This is a critical area to pay attention to as it is an option offered on every question, selected by up to 4% of participants on key questions such as “If the presidential election were today, how would you vote?” (cite Fox article).

>>>>

## Implications and Potential Weaknesses {#sec-disc-imp}

Our model predicting the U.S. election carries important social and political implications. Accurate forecasts can shape voter perceptions, potentially affecting turnout by leading some individuals to feel their votes are less crucial if a particular outcome seems likely. Additionally, these predictions inform campaign strategies, prompting candidates to focus resources on undecided voters in critical swing states, which play a pivotal role in tight elections like the upcoming 2024 race. For instance, our model anticipates that key swing states such as Pennsylvania, Michigan, and Wisconsin will award their electoral votes to Kamala Harris, contributing to her victory. However, the forecast indicates that Harris is expected to win by narrow margins of approximately 1.98%, 1.47%, and 1.46% in these states, respectively.

Our model offers valuable insights into the landscape of the current U.S. election; however, it does face a few limitations. One of the primary challenges stems from gaps in our dataset, specifically the absence of polling data for several states. When the polling data was collected and compiled by @538, it did not comprehensively include every state, leading to missing data points that affect our predictions. Our model relies on the electoral college system, where each state’s popular vote determines its allocation of electoral votes, ultimately deciding the election’s outcome. Missing data from certain states introduces an incomplete representation of the voting scene, affecting the accuracy of our predictions. Each state often exhibits unique voting behaviors and excluding certain states can introduce a bias in our model. This is especially problematic if the omitted states are swing states or tend to have historically unpredictable voting patterns, as these states can play a decisive role in election outcomes. Moreover, missing data for larger states further exacerbates the issue by potentially leading to a miscalculation of the Electoral College outcome. States with significant electoral votes carry substantial weight, and without accurate polling information from them, our model may struggle to forecast the election winner effectively.

Another limitation of our model is that we stopped pulling new poll data from @538 as of October 19, 2024. Consequently, the model does not account for any last-minute events that could have impacted voter opinions from then until election day. This could potentially lead to misleading predictions if there has been a significant change in public sentiment. Furthermore, last-minute shifts in public opinion could result in different voter turnout than initially anticipated or introduce polling response biases, which predictive models may struggle to capture if they are based primarily on earlier data.

## Next Steps {#sec-disc-next}

Consequently, our forecast suggests that candidates should concentrate their efforts on swing states, as increased campaigning in these areas could significantly impact voter turnout and the popular vote. Additionally, enlisting high-profile figures or celebrities, like Elon Musk or LeBron James—who have recently endorsed Trump and Harris, respectively—could enhance their appeal in these swing states. These endorsements would help candidates reach diverse and niche demographics, broadening their voter base and potentially swaying undecided voters.

## Betting Markets {#sec-disc-bet}

@@

=======
## Analysis into a Pollster's Methodology

One particular poll within our data sample is one conducted between October 11-14 by Beacon Research and Shaw & Company Research [@fox]. Targeting the sample frame of registered American voters, this poll collected data on 1,110 participants’ responses on 51 questions through either live phonecall interviews or an online survey. The sample was recruited by applying the probability proportionate to size method on the nationwide voter file of registered voters’ phone numbers [@fox]. This sampling method is a type of probability sampling in which each unit's chance of being included is dependent upon their size, a measure that is based off of some characteristic that is known about every single unit and often related to the main variable of interest. By giving a higher probability of inclusion to units of greater importance based on its size during sampling, a more accurate estimate can be obtained [@Latpate2021]. In the case of this poll, the number of voters per state region was used to determine the proportion of individuals contacted [@fox]. This means a greater number of residents in larger states, such as California, would be contacted compared to smaller states, such as Hawaii. This method of sampling ensures that the voices represented by poll results would match the ones involved with the real election as much as possible.

The key finding from this poll is that 48% of participants favored Harris while 50% favored Trump, with the lead being consistent in the larger sample of registered voters and smaller subsample of likely voters. Despite this, responses also indicate that Harris is narrowly leading in seven swing states, meaning that Democrats could potentially win by electoral college while losing the popular vote  [@fox]. This report also notes that current results are Trump’s highest approval ratings since Biden dropped out of the race, while support for Harris is at its lowest.

One strength of this questionnaire is that information regarding the respondent’s confidence level and commitment level was also collected. With questions such as “How often do you make a point to read or listen to the news?” and “Are you certain to support that candidate, or do you think you may change your mind and support someone else?”, this poll is able to gain a more indepth view of how easily these participants may be swayed in the time between the polling and the real election  [@fox]. As well, probabilistic statistical models based on past voting history, interest in current election, age, education, race, ethnicity, church attendance, and marital status were used to predict the respondents’ likeliness to vote; interestingly, results found that the results for the full sample vs subsample of 870 likely voters varied by a ±3%, which can be quite significant given how tight the race is currently  [@fox].

On the other hand, one potential weakness of this poll is that it was sponsored by Fox News, a news site that has historically been Republican-leaning. Though from the reported methodology alone, no evident biasing of the pollster can be observed, this potential source of biased reporting must be noted as it can affect the type of analyses that occurred and which key findings are the focus of news reports. As well, another potential issue is the lack of indication regarding how non-responses of “Don’t know” were handled during analysis  [@fox]. This is a critical area to pay attention to as it is an option offered on every question, selected by up to 4% of participants on key questions such as “If the presidential election were today, how would you vote?” [@fox].
